Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-16 paper/paper.tex
paper: add groundtruth feature effects appendix

The need for interpretability in ML stems from multiple factors: building trust in
model predictions~\cite{ribeiro_why_2016,teach_analysis_1981},  identifying and mitigating
potential biases~\cite{guidotti_survey_2019}, addressing fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensuring compliance with regulations like the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, develping a
wide variety of methods to explain ML models. This includes both model-specific and
model-agnostic approaches ranging from local feature attributions to global
feature importances and effects\footnote{For comprehensive overviews of Explainable AI
    methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.

@@

However, to the best of our knowledge, there exist no such systematic studies
for feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}. Current literature
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
predominantly uses training data without explicit justification, while
practitioners also advocate for using unseen validation data (for details, see
\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
choice include potential biases arising from overfitting, the dataset size, and
computational constraints. Although PDP and ALE are not based on generalization
error like PFI, it is not studied if and how they are affected by overfitting
when estimated on training data. In addition, larger datasets may improve the
feature effect estimates, but also increase computation times.\\

@@

\noindent These contributions have several important implications: First, our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which data to choose for feature effect estimation, helping to
understand potential implications of their choices. Additionally, our framework
for evaluating feature effects and systematic collection of test functions
provide a foundation for future research in Interpretable ML, with the latter
specifically facilitating test function choice for simulation studies.\\

@@

>>>>

The need for interpretability in ML stems from multiple factors: building trust in
model predictions~\cite{ribeiro_why_2016,teach_analysis_1981},  identifying and mitigating
potential biases~\cite{guidotti_survey_2019}, addressing fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensuring compliance with regulations like the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, developing a
wide variety of methods to explain ML models. This includes both model-specific and
model-agnostic approaches ranging from local feature attributions to global
feature importances and effects\footnote{For comprehensive overviews of Explainable AI
    methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.

@@

However, to the best of our knowledge, there exist no such systematic studies
for feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}. Current literature
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
predominantly uses training data without explicit justification, while
practitioners also advocate for using unseen validation data (for details, see
\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
choice include potential biases arising from overfitting, the dataset size, and
computational constraints. Although PDP and ALE are not based on generalization
error like PFI, it is not studied if and how they are affected by overfitting
when estimated on training data. In addition, larger datasets may improve the
feature effect estimates but also increase computation times.\\

@@

\noindent These contributions have several important implications: First, our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which data to choose for feature effect estimation, helping to
understand potential implications of their choices. Additionally, our framework
for evaluating feature effects and our systematic collection of test functions
provide a foundation for future research in Interpretable ML, with the latter
specifically facilitating test function choice for simulation studies.\\

@@

\clearpage
\section{Derivation and Visualization of Analytical Ground Truth Feature Effects}\label{app:ground-truth-feature-effects}
In the following, we derive the analytical ground truth PDPs and ALEs for the
considered datasets. In addition, we provide visualizations of the derived
feature effects and compare them to the estimated feature effects as gut check.

\begin{figure*}[h!]
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/pdp_groundtruth_comparison.png}
        \caption{PDP ground truth of SimpleNormalCorrelated}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/ale_groundtruth_comparison.png}
        \caption{ALE ground truth of SimpleNormalCorrelated}
    \end{subfigure}
    \caption{Ground truth feature effects for SimpleNormalCorrelated dataset. The green line
    represents the analytically derived ground truth, the orange line the estimated feature
    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
    are centered that the average effect at the quantile-based grid points is 0.
    Since $x_3$ and $x_4$, are dummy variables, they have a constant effect of 0 and are
    thus omitted.}
    \label{fig:snc-groundtruth}  % chktex 24
\end{figure*}

\noindent For SimpleNormalCorrelated, we use normally distributed features with $\mu_1 = 0$,
$\mu_2 = 0$, $\sigma_1 = 1$, $\sigma_2 = 1$, and correlation
$\rho = 0.9$. We denote
arbitrary constants with $C$ and $C^\prime$, respectively.

\begin{align*}
    PDP_{f,1}(x_1) &= \mathbb{E}_{X_2}[x_1 + \frac{1}{2}X_2^2 + x_1 X_2] \\
                 &= x_1 + \frac{1}{2} \mathbb{E}_{X_2}[X_2^2] + x_1 \mathbb{E}_{X_2}[X_2] \\
                 &= x_1 + \frac{1}{2} (Var_{X_2}(X_2) + \mathbb{E}_{X_2}[X_2]^2) + x_1 \mathbb{E}_{X_2}[X_2] \\
                 &= x_1 + \frac{1}{2} (\sigma_2^2 + \mu_2^2) + x_1 \mu_2 \\
                 &= x_1 + \frac{1}{2} \\
                 &= x_1 + C_1
\end{align*}

\begin{align*}
    PDP_{f,2}(x_2) &= \mathbb{E}_{X_1}[X_1 + \frac{1}{2}x_2^2 + X_1 x_2] \\
                    &= \mathbb{E}_{X_1}[X_1] + \frac{1}{2}x_2^2 + \mathbb{E}_{X_1}[X_1] x_2 \\
                    &= \mu_1 + \frac{1}{2}x_2^2 + \mu_1 x_2 \\
                    &= \frac{1}{2}x_2^2
\end{align*}

\begin{align*}
    ALE_{f,1}(x_1) &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[\frac{\partial}{\partial X_1} X_1 + 0.5 X_2^2 + X_1X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
    &(\text{using } X_2 = \beta_{2|1}(X_1-\mu_1) + \varepsilon_1 + \mu_2\text{, where } \beta_{2|1}=\rho \frac{\sigma_2}{\sigma_1}) \\
    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + \beta_{2|1}(X_1-\mu_1) + + \varepsilon_1 + \mu_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
    &= \int_{x_{min,1}}^{x_1} 1 + \beta_{2|1}(z_1-\mu_1) + \mu_2 \,dz_1 - C_1 \\
    &= [z_1 + \frac{1}{2}\beta_{2|1}(z_1-\mu_1)^2 + \mu_2 z_1]_{x_{min,1}}^{x_1} - C_1 \\
    &= x_1 + \frac{\beta_{2|1}}{2}(x_1-\mu_1)^2 + \mu_2 x_1 \\
    & \quad - x_{min,1} - \frac{\beta_{2|1}}{2}(x_{min,1}-\mu_1)^2 - \mu_2 x_{min,1} - C_1 \\
    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot ((x_1-\mu_1)^2 - (x_{min,1}-\mu_1)^2) - C \\
    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot (x_1^2 - 2\mu_1 x_1 - x_{min,1}^2 + 2\mu_1 x_{min,1}) - C \\
    &= (1 + \mu_2) x_1 - (1 + \mu_2) x_{min,1} + \frac{\beta_{2|1}}{2} x_1^2 - \frac{\beta_{2|1}}{2} x_{min,1}^2 \\
    & \quad - \mu_1 \beta_{2|1} x_1 - \mu_1 \beta_{2|1}  x_{min,1} - C_1 \\
    &= \frac{\beta_{2|1}}{2} x_1^2 + (1 + \mu_2 - \mu_1 \beta_{2|1}) x_1 - C_1^\prime \\
    &= 0.45 x_1^2 + x_1 - C_1^\prime
\end{align*}
\\

\noindent With anologous calculations, we get:

\begin{align*}
    ALE_{f,2}(x_2) &= (\frac{1}{2} + \mu_1 + \frac{\beta_{1|2}}{2}) x_2^2 - \beta_{1|2} \mu_2 x_2 - C_2^\prime \\
    &= 0.95 x_2^2 - C_2^\prime
\end{align*}

\subsection{Friedman1 Dataset}

For Friedman1, we use independent uniformly distributed features between 0
and 1. Thus, all features $S$ have $\mu_S = \mathbb{E}[X_S]=\frac{a+b}{2}=0.5$ and
$\sigma_S^2 = Var[X]=\frac{(b-a)^2}{12}=\frac{1}{12}$.
Si denotes the sine integral function, $C_S$ and $C_S^\prime$ are arbitrary constants.

\begin{align*}
    PDP_{f,1}(x_1) &= \mathbb{E}_{-X_1}[10\sin(\pi x_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 + 5X_5] \\
    &= 10 \mathbb{E}_{X_2}[\sin(\pi x_1X_2)] + 20(\mathbb{E}_{X_3}[X_3^2] - \mathbb{E}_{X_3}[X_3] + \frac{1}{4}) + 10 \mathbb{E}_{X_4}[X_4] \\
    &\quad + 5 \mathbb{E}_{X_5}[X_5]\\
    &= 10 \cdot \int_{a}^{b} \sin(\pi x_1x_2)dx_2 + 20\cdot(\mu_3^2 + \sigma_3^2 - \mu_3 + \frac{1}{4}) + 10\mu_4 + 5\mu_5 \\
    &= 10 \cdot \frac{-\cos(\pi bx_1) + \cos(\pi ax_1)}{\pi x_1} + 20(\mu_3^2 + \sigma_3^2 - \mu_3) + 5 + 10\mu_4 + 5\mu_5 \\
    &= 10 \cdot \frac{-\cos(\pi x_1) + 1}{\pi x_1} + C_1 \\
\end{align*}

\noindent Analogously, we get for $x_2$:

\begin{align*}
    PDP_{f,2}(x_2) &= 10 \cdot \frac{-\cos(\pi x_2) + 1}{\pi x_2} + C_2
\end{align*}

\begin{align*}
    PDP_{f,3}(x_3) &= \mathbb{E}_{-X_3}[10\sin(\pi X_1X_2) + 20(x_3-\frac{1}{2})^2 + 10X_4 + 5X_4] \\
    &= 10\int_{a_1}^{b_1}\int_{a_2}^{b_2}\sin(\pi x_1x_2)dx_2dx_1 + 20(x_3-\frac{1}{2})^2 + 10\mu_4 + 5\mu_5 \\
    &= \frac{10}{\pi}[Si(\pi b_1a_2) - Si(\pi a_1a_2) - Si(\pi b_1b_2) + Si(\pi a_1b_2)] \\
    &\quad + 20x_3^2 - 20x_3 + 5 + 10\mu_4 + 5\mu_5 \\
    &= 20x_3^2 - 20x_3 + C_3 \\
\end{align*}

\noindent Since Si is not defined at 0, we use a small $\epsilon$ for
computational stability.\newline
Analogously, we get for $x_4$, $x_5$:

\begin{align*}
    PDP_{f,4}(x_4) &= 10x_4 + C_4
\end{align*}

\begin{align*}
    PDP_{f,2}(x_2) &= 5x_5 + C_5
\end{align*}

\begin{align*}
    ALE_{f,1}(x_1) &= \dots
\end{align*}

\begin{align*}
    ALE_{f,2}(x_2) &= \dots
\end{align*}

\begin{align*}
    ALE_{f,3}(x_3) &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\Bigg[\frac{\partial}{\partial x_3}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
    &\qquad\qquad\qquad\qquad\qquad \Bigg. + 5X_5) \big|X_3=z_3\Bigg]dz_3 - C_3 \\
    &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\left[40z_3-20\right]dz_3 - C_3 \\
    &= \int_{x_{3,\min}}^{x_3} 40z_3-20dz_3 - C_3 \\
    &= \left[20z_3^2 - 20z_3\right]_{x_{3,\min}}^{x_3} - C_3 \\
    &= 20x_3^2 - 20x_3 - 20x_{3,\min}^2 + 20x_{3,\min} - C_3 \\
    &= 20x_3^2 - 20x_3 - C_3^\prime
\end{align*}

\begin{align*}
    ALE_{f,4}(x_4) &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-X_4|X_4}\Bigg[\frac{\partial}{\partial x_4}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
    &\qquad\qquad\qquad\qquad\qquad \Bigg. {} + 5X_5) \big|X_4=z_4\Bigg]dz_4 - C_4 \\
    &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-x_4|x_4}\left[10\right]dz_4 - C_4 \\
    &= 10x_4 - 10x_{4,\min} - C_4 \\
    &= 10x_4 - C_4^\prime
\end{align*}

\noindent Analogously, we get for $x_5$:

\begin{align*}
    ALE_{f,5}(x_5) &= 5x_5 - C_5^\prime
\end{align*}

\begin{figure*}[h!]
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/pdp_groundtruth_comparison.png}
        \caption{PDP ground truth of Friedman1}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/ale_groundtruth_comparison.png}
        \caption{ALE ground truth of Friedman1}
    \end{subfigure}
    \caption{Ground truth feature effects for Friedman1 dataset. The green line
    represents the analytically derived ground truth, the orange line the estimated feature
    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
    are centered that the average effect at the quantile-based grid points is 0.
    Since $x_6$ and $x_7$, are dummy variables, they have a constant effect of 0 and are
    thus omitted.}
    \label{fig:friedman1-groundtruth}  % chktex 24
\end{figure*}

\subsection{FeynmanI.29.16 Dataset}

For the FeynmanI.29.16 dataset, it is not possible to derive the ground truth
feature effects analytically. Nonetheless, we give a visualization of the
estimated ground truth feature effects for illustration purposes in
\textbf{Fig.~\ref{fig:feynman-groundtruth}}.

\begin{figure*}[h!]
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/pdp_groundtruth_comparison.png}
        \caption{PDP ground truth of FeynmanI.29.16}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/ale_groundtruth_comparison.png}
        \caption{ALE ground truth of FeynmanI.29.16}
    \end{subfigure}
    \caption{Ground truth feature effects for FeynmanI.29.16 dataset. The green line
    represents the analytically derived ground truth, the orange line the estimated feature
    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
    are centered that the average effect at the quantile-based grid points is 0.
    Since $d_1$ and $d_2$, are dummy variables, they have a constant effect of 0 and are
    thus omitted.}
    \label{fig:feynman-groundtruth}  % chktex 24
\end{figure*}
